Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-21 doc/mc-paper.tex
Minor edits

\section{Implementations}
We have implemented versions of the three programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}.  The proof generator is the same in both
cases, since it need not be trusted.
Our {\em verified}
versions of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our {\em prototype} versions are written in C
(checker) and Python (ring evaluator).  These are currently faster and
more scalable, but we anticipate their need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

@@

Our prototype ring evaluator can perform both standard and weighted
model counting.  It performs arithmetic over a subset of the rationals
we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
to have arbitrary range.  Allowing scaling by powers of 2 enables the
density computation and rescaling required for standard model
counting.  Allowing scaling by powers of both 2 and 5 enables exact
decimal arithmetic, handling the weights used in the weighted model
counting competitions.  For one of the weighted benchmarks, it
generated a result with 260,909 decimal digits.

@@

We provide only a summary of our experiments here.  A more complete
description is provided in Appendix~\ref{app:experiments}\@. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We then ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graphs generated by \dfour{}.

>>>>

\section{Implementations}
We have implemented versions of the three programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
benchmark problems are available at \url{https://github.com/rebryant/crat}}.  The proof generator is the same in both
cases, since it need not be trusted.
Our {\em verified}
versions of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our {\em prototype} versions are written in C
(checker) and Python (ring evaluator).  These are currently faster and
more scalable, but we anticipate their need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

@@

Our prototype ring evaluator can perform both standard and weighted
model counting.  It performs arithmetic over a subset of the rationals
we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
to have arbitrary range.  Allowing scaling by powers of 2 enables the
density computation and rescaling required for standard model
counting.  Allowing scaling by powers of both 2 and 5 enables exact
decimal arithmetic, handling the weights used in the weighted model
counting competitions.  To give a sense of scale, the counter
generated a result with 260,909 decimal digits
for one of the weighted benchmarks.

@@

We summarize the results of our experiments here.  
Appendix~\ref{app:experiments} provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graphs generated by \dfour{}.
